Understanding the European Football by Jorge Saldivar
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.
library(ggplot2)
library(dplyr)
library(ggthemes)
library(knitr)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
dictionary = read.csv('dictionary.csv', stringsAsFactors = F, header = T)
events = read.csv('events.csv', stringsAsFactors = F, header = T)
# Using the given dictionary, replace codes for their corresponding values
for (current_category in unique(dictionary$category)) {
  if (current_category %in% names(events)) {
    data_category = filter(dictionary, category==current_category)
    for (code_value in data_category$code) {
      events[!is.na(events[,current_category])&events[,current_category]==code_value,current_category] = data_category[data_category$code==code_value,'description'] 
    }
  }
}
```

In this study, I analyze football (soccer in US) events through a dataset composed of `r nrow(events)` events collected from 9,074 european games. The dataset is open source and available to download at https://www.kaggle.com/secareanualin/football-events/data. Details of the events are represented through `r ncol(events)` features, including the type of event (11 types, including foul, corner, offside, and red card), the minute in which the event happens, the side of the team involved in the event (home or away), the location of the event in the pitch (e.g., defensive half, attacking half, penalty spot), the game situation in which the event happens (e.g., open play, corner, free kick). If the event is a shot, the dataset provides also information of the shot place (e.g., bottom left corner, top left corner, center of the goal), the shot outcome (e.g., on target, off target, blocked), the part of the body involved in the shot (i.e., left foot, right foot, head), the method used to assist the shoter (e.g., pass, cross, headed pass), and a boolean variable that indicates whether the shot ends up as a goal. There are complementary data about the events such as the team that generates the event or the players involved.

# Univariate Analysis

This analysis focused on 11 of the features, namely: event type, time, side, location, situation, shot place, shot outcome, body part, assist method, and team generator. No new variables were created for the study but to facilitate the analysis, the codes of the categorical variables are replaced with their textual values. The mapping between categories and values is recorded in a dictionary file that can be download together with the dataset from https://www.kaggle.com/secareanualin/football-events/data. As an example, next, it is shown five rows of the datasset.

```{r, echo=FALSE}
cols_to_show = c('event_type', 'time' ,'event_team', 'side', 'location', 'situation', 'shot_place', 'shot_outcome', 'bodypart', 'assist_method', 'is_goal')
kable(head(events[11:15,cols_to_show]))
```

# Univariate Plots Section

### Event Types

```{r echo=FALSE, Univariate_Plots}
# order levels in decreasing order
event_types = factor(events$event_type, levels=names(sort(table(events$event_type), decreasing=TRUE)))
tevents = data.frame(table(event_types))
colnames(tevents) = c('event_type', 'count')
ggplot(aes(x=event_type, y=count, fill=event_type), data=tevents) +
  geom_bar(colour="black", stat = "identity") +
  geom_text(aes(label=count), position=position_dodge(width=0.9), vjust=-0.50) +
  guides(fill=FALSE) +
  theme_economist() +
  xlab('Event Type') +
  ylab('Count') +
  theme(axis.text.x  = element_text(angle=90, vjust=0.5, size=10)) +
  ylim(0, 250000)
```

The figure above help us to have a first overview on the primary characteristics of the European football. Fouls and free kicks are the two most common events in games, followed by goal attempts --- they represent `r round(100*(sum(tevents[1:3,'count'])/nrow(events)),1)`% of the events in the dataset. The large occurrence of fouls and free kicks demonstrates that games are apparently not fluent but frenquently interrupted by ilegal plays and the resulting direct or indirect kicks. It is interesting to notice that even when fouls are among the most common events, games are characterized by fair plays considering the low frequency of yellow and red cards (`r round(100*(sum(tevents[c(7,10),'count'])/nrow(events)),1)`% of the total events). The figure shows that corners are not that common, which seems to indicate that goal attempts come more from open plays (e.g., give and go) than through set pieces. The low occurrence of penalties (`r round(100*(tevents[9,'count']/nrow(events)),1)`% of the total events) indicates that fouls are committed outside the goal area.

### Time

```{r echo=FALSE, Univariate_PlotsT}
ggplot(aes(x=time), data=events) +
  geom_histogram(binwidth = 1) +
  theme_economist() +
  xlab('Time') +
  ylab('Count') +
  scale_x_continuous(breaks=seq(0,100,5))
```


The figure above shows that there is a slight but consistent increment in the number of events from the first minute to the end of the game. The final minutes of each half (45th and 90th) concentrate a large amount of events.

### Side

```{r echo=FALSE, Univariate_Plots2}
# order levels in decreasing order
side = factor(events$side, levels=names(sort(table(events$side), decreasing=TRUE)))
tside = data.frame(table(side))
colnames(tside) = c('side', 'count')
ggplot(aes(x=side, y=count, fill=side), data=tside) +
  geom_bar(colour="black", stat = "identity") +
  geom_text(aes(label=count), position=position_dodge(width=0.9), vjust=-0.50) +
  guides(fill=FALSE) +
  theme_economist() +
  xlab('Side') +
  ylab('Count') +
  theme(axis.text.x  = element_text(vjust=0.5, size=10)) +
  ylim(0, 500000)
```

The amount events per side are balanace between those related with home teams (`r round(100*(tside[1,'count']/nrow(events)),1)`%) and the ones belonged to away teams.

### Location

```{r echo=FALSE, Univariate_Plots3}
# order levels in decreasing order
location = factor(events$location, levels=names(sort(table(events$location), decreasing=TRUE)))
tlocation = data.frame(table(location, useNA="ifany"))
colnames(tlocation) = c('location', 'count')
ggplot(aes(x=location, y=count, fill=location), data=tlocation[1:19,]) +
  geom_bar(colour="black", stat = "identity") +
  geom_text(aes(label=count), position=position_dodge(width=0.9), vjust=-0.50, size=3) +
  guides(fill=FALSE) +
  theme_economist() +
  xlab('Location') +
  ylab('Count') +
  theme(axis.text.x  = element_text(angle=90, vjust=0.5, size=10)) +
  ylim(0, 150000)
```

Less than half of the records (`r round(100*(sum(tlocation[1:19,'count'])/nrow(events)),1)`%) have registered the location of pitch where the event happened. The records with blank location are not included in the calculation of the figure above. Some results in this figure might complement the insights obtained in the analysis of the event types. For example, fouls might occurre in the majority of cases in the defensive half and probably outside of the box. Further analyses are required to better understand relationship between variables.

### Game situation

```{r echo=FALSE, Univariate_Plots4}
# order levels in decreasing order
situation = factor(events$situation, levels=names(sort(table(events$situation), decreasing=TRUE)))
tsituation = data.frame(table(situation, useNA="ifany"))
colnames(tsituation) = c('situation', 'count')
ggplot(aes(x=situation, y=count, fill=situation), data=tsituation[1:4,]) +
  geom_bar(colour="black", stat = "identity") +
  geom_text(aes(label=count), position=position_dodge(width=0.9), vjust=-0.50, size=5) +
  guides(fill=FALSE) +
  theme_economist() +
  xlab('Game situation') +
  ylab('Count') +
  theme(axis.text.x  = element_text(vjust=0.5, size=10)) +
  ylim(0, 200000)
```

Only `r round(100*(sum(tsituation[1:4,'count'])/nrow(events)),1)`% of the records are equipped with the situation that triggered the event. Records without mention of the game situation are not considered to draw the figure above that shows that in most cases events started from open plays. To a lesser extend, corners, set pieces, and free kicks caused events as well.

### Shot place

```{r echo=FALSE, Univariate_Plots5}
# order levels in decreasing order
shot_place = factor(events$shot_place, levels=names(sort(table(events$shot_place), decreasing=TRUE)))
tshot_place = data.frame(table(shot_place, useNA="ifany"))
colnames(tshot_place) = c('shot_place', 'count')
ggplot(aes(x=shot_place, y=count, fill=shot_place), data=tshot_place[1:13,]) +
  geom_bar(colour="black", stat = "identity") +
  geom_text(aes(label=count), position=position_dodge(width=0.9), vjust=-0.50, size=4) +
  guides(fill=FALSE) +
  theme_economist() +
  xlab('Shot place') +
  ylab('Count') +
  theme(axis.text.x  = element_text(angle=90, vjust=0.5, size=10)) +
  ylim(0, 60000)
```

Out of the `r nrow(events[events$event_type=='Attempt',])` attempts, `r sum(tshot_place[1:13,'count'])` (`r round(100*(sum(tshot_place[1:13,'count'])/nrow(events[events$event_type=='Attempt',])),1)`%) register where the shot ended up. The records without information about shot place are excluded from figure above (`r tshot_place[14,'count']` records). Most of the shots were blocked, of the remaining a large number of shots missed to one of the sides (left or right) or go to the center of the goal.

### Shot outcome

```{r echo=FALSE, Univariate_Plots6}
# order levels in decreasing order
shot_outcome = factor(events$shot_outcome, levels=names(sort(table(events$shot_outcome), decreasing=TRUE)))
tshot_outcome = data.frame(table(shot_outcome, useNA="ifany"))
colnames(tshot_outcome) = c('shot_outcome', 'count')
ggplot(aes(x=shot_outcome, y=count, fill=shot_outcome), data=tshot_outcome[1:4,]) +
  geom_bar(colour="black", stat = "identity") +
  geom_text(aes(label=count), position=position_dodge(width=0.9), vjust=-0.50, size=4) +
  guides(fill=FALSE) +
  theme_economist() +
  xlab('Shot outcome') +
  ylab('Count') +
  theme(axis.text.x  = element_text(vjust=0.5, size=10)) +
  ylim(0, 100000)
```

Out of the `r nrow(events[events$event_type=='Attempt',])` attempts, `r sum(tshot_outcome[1:4,'count'])` (`r round(100*(sum(tshot_outcome[1:4,'count'])/nrow(events[events$event_type=='Attempt',])),1)`%) register the shot outcome. The figure above does not include the records that do not report the shot outcome (`r tshot_outcome[5,'count']` records). In the figure above we can see that shoters have a probability of `r round((tshot_outcome[2,'count']/sum(tshot_outcome[1:4,'count']))*100,1) `% of hitting the target.

### Body part

```{r echo=FALSE, Univariate_Plots7}
# order levels in decreasing order
body_part = factor(events$bodypart, levels=names(sort(table(events$bodypart), decreasing=TRUE)))
tbody_part = data.frame(table(body_part, useNA="ifany"))
colnames(tbody_part) = c('body_part', 'count')
ggplot(aes(x=body_part, y=count, fill=body_part), data=tbody_part[1:3,]) +
  geom_bar(colour="black", stat = "identity") +
  geom_text(aes(label=count), position=position_dodge(width=0.9), vjust=-0.50, size=4) +
  guides(fill=FALSE) +
  theme_economist() +
  xlab('Body Part') +
  ylab('Count') +
  theme(axis.text.x  = element_text(vjust=0.5, size=10)) +
  ylim(0, 130000)
```

All of the events of type *attempt* are equipped with information about the body part involved in the event. In the figure above we can see that the majority of attempts (`r round(100*(tbody_part[1,'count']/nrow(events[events$event_type=='Attempt',])),1)`% of the attempts) were performed using the right feet.

### Assist method

```{r echo=FALSE, Univariate_Plots8}
# order levels in decreasing order
assist_method = factor(events$assist_method, levels=names(sort(table(events$assist_method), decreasing=TRUE)))
tassist_method = data.frame(table(assist_method, useNA="ifany"))
colnames(tassist_method) = c('assist_method', 'count')
ggplot(aes(x=assist_method, y=count, fill=assist_method), data=tassist_method) +
  geom_bar(colour="black", stat = "identity") +
  geom_text(aes(label=count), position=position_dodge(width=0.9), vjust=-0.50, size=4) +
  guides(fill=FALSE) +
  theme_economist() +
  xlab('Assist Method') +
  ylab('Count') +
  theme(axis.text.x  = element_text(vjust=0.5, size=10)) +
  ylim(0, 800000)
```

Passes represent the most common assist method followed by crosses. It is interesting to see that crosses are not frequently used in the top leagues of Europe as a method to assist strikers. This can be different in other parts of the world, like south-america, where cross is very a common assistance method.

### Was it goal?

```{r echo=FALSE, Univariate_Plots9}
# order levels in decreasing order
t_goal = data.frame(table(events$is_goal, useNA="ifany"))
colnames(t_goal) = c('code', 'count')
t_goal$code = as.numeric(t_goal$code)
t_goal[t_goal$code==1,'code'] = 'No'
t_goal[t_goal$code==2,'code'] = 'Yes'
ggplot(aes(x=code, y=count, fill=code), data=t_goal) +
  geom_bar(colour="black", stat = "identity") +
  geom_text(aes(label=count), position=position_dodge(width=0.9), vjust=-0.50, size=4) +
  guides(fill=FALSE) +
  theme_economist() +
  xlab('Goal') +
  ylab('Count') +
  theme(axis.text.x  = element_text(vjust=0.5, size=10)) +
  ylim(0, 920000)
```

The vast majority of events were not goals. Out of the `r nrow(events[events$event_type=='Attempt',])` attempts, `r t_goal[t_goal$code=='Yes','count']`  were goals. An effectiveness of `r round((t_goal[t_goal$code=='Yes','count']/nrow(events[events$event_type=='Attempt',]))*100,1)` %.

## Event team generator

```{r echo=FALSE, Univariate_Plots10}
# order levels in decreasing order
event_teams = factor(events$event_team, levels=names(sort(table(events$event_team), decreasing=TRUE)))
t_teams = data.frame(table(event_teams, useNA="ifany"))
colnames(t_teams) = c('team', 'count')
ggplot(aes(x=team, y=count, fill=team), data=t_teams[1:20,]) +
  geom_bar(colour="black", stat = "identity") +
  geom_text(aes(label=count), position=position_dodge(width=0.9), vjust=-0.50, size=2) +
  guides(fill=FALSE) +
  theme_economist() +
  xlab('Top-20 teams that generate more events') +
  ylab('Number of events') +
  theme(axis.text.x  = element_text(angle=90,vjust=0.5, size=10)) +
  ylim(0, 13000)
```

The dataset contains information about events generated by `r nrow(t_teams)` from Italy, Germany, France, Spain, and England. Taking a closer look at  the top-20 teams that generate more events, it can be seen that half of them belong to the Serie A (Italy), namely: Juventus, Fiorentina, AC Milan, AS Roma, Genova, Lazio, Atalanta, Napoli, Chievo, and Inter. Out of the rest, 5 teams are Spanish and 5 play in the Ligue 1 in France.

# Bivariate Plots Section

> **Tip**: Based on what you saw in the univariate plots, what relationships
between variables might be interesting to look at in this section? Don't limit
yourself to relationships between a main output feature and one of the
supporting variables. Try to look at relationships between supporting variables
as well.

```{r echo=FALSE, Bivariate_Plots}

```

# Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

### What was the strongest relationship you found?


# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!